Acceleration of Derivative Calculations with Application to Radial Basis Function -- Finite-Differences on the Intel MIC Architecture

### 2013-10-19 ieee_paper.tex

\begin{document}
\title{Acceleration of Derivative Calculations with Application to Radial Basis Function --
Finite-Differences on the MIC Architecture}
%\title{Sparse Matrix Vector Multiplication with Multiple vectors and
  %Multiple Matrices on the MIC Architecture \todo{? more application oriented''}}

@@

In this paper, we consider the case of local stencils with the number
of nodes at each point and encode the sparse matrix in ELLPACK
format. We increase the number of operations relative to memory
bandwidth by calculating four derivatives of four different functions,
or 16 different derivatives. We demonstrate a novel implementation on
the MIC architecture, taking into account its advanced swizzling and
channel interchange features. We present benchmarks on a real data
set that show an almost eightfold increase in speed compared to
efficient implementations of a single derivative, reaching a
performance of almost 140~Gflop/s. We explain the results through
consideration of operation count versus memory bandwidth.
\end{abstract}

@@

We discuss how the implementation of the v4m4 kernel works using
vectorial instructions. The code is given for reference in
Figure~\ref{code:mat_mul}. (We will show in Section~\ref{sec:expe}
that an implementation that relies on compiler vectorization does not lead
to desirable performance, which confirms the results
of~\cite{Saule13-ARXIV}.) The registers in the MIC architecture are
512 bits wide and can store 16 floats or integers. (In comparison to
OpenCL or CUDA, one can think of these registers as a warp.) The end
goal is to load and format the nonzero values $A_h^1,A_h^2,A_h^3,A_h^4$
and the vector entries $x_m^1,x_m^2,x_m^3,x_m^4$ in two vector registers and 
apply Fused Multiply-Add on them, which is depicted in the {\tt accu} line of 
Figure~\ref{code:mat_mul}. Bringing the data into the vector
register as depicted in the figure is the difficult component of our 
proposed algorith, which we proceed to explain further. 
A thread will perform the multiplications one row at a time,
and parallelism is achieved by giving blocks of rows to each core
using an OpenMP construct.

The core of the technique is the use of swizzling and permutation
features of the vectors in the Intel MIC architecture. A vector of 512
bits is composed of four lanes (sometimes called channels) of 128
bits, which are made of four segments of 32 bits. A swizzling
operation allows the reorder of the elements within each lane (see how
{\tt mat} is extracted from {\tt mat\_ent} in
Figure~\ref{code:mat_mul}). While a permutation reorders entire lanes,
without changing their contents (see for instance how {\tt vect} is
generated from {\tt all\_vect} in Figure~\ref{code:mat_mul}).  Both
permutation and swizzling support common reordering and broadcast. In
the code, the ``permute'' function needs to use the {\tt
  \_mm512\_permute4f128\_epi32} instruction which is only defined on
integers and therefore in place typecasts are necessary to convert to
a vector to/from the integer type.  (Note that we prefer swizzling to
permutation if possible because we believe it is cheaper. Many
instructions have swizzling capability, while permuation is its own
instruction. %\todo{do we need a reference to intel Xeon phi instruction set?})

Every operation that loads data from memory into a register
transfers 512 bits, or 16 floats at a time. A single nonzero
in the matrix will cause 16 operations to be performed, but it only
uses 4 floats from the matrix and 4 floats from the {\tt x} vector. So
we would like to execute a single instruction to load the data for four 
nonzeros in the matrix, and another instruction to load the data for
four nonzeros from the vectors.  
%representing 16 floats from the matrix and 16 floats from the vectors at once. 
(The number of nonzeros on a row in our application is
always a multiple of 4; if it weren't, we would add some explicit
zeroes in the data structure for padding.)
\todo{Erik, I rewrote the paragraph. Please reread.}

@@

Figure~\ref{fig:bandwidth} gives some benchmark that computes the
bandwidth achieved when reading and/or writing large arrays on both
the MIC card and the CPU architecture. The read benchmark performs a
simple sum of the array. The write benchmark sets all the elemnts of
the array to zero, while the read/write benchmarks copies the array into
another one. We also investigate the difference between a straight C++
implementation (compiled with -O3 with proper tagging to inform the
compiler of memory alignments) and a simple implementation with
explicit use of the vector registers.

When using the MIC card (Figure~\ref{fig:band_rw_MIC}), the {\tt
  read\_cpp} implementation reaches a performance of 168~GB/s
 which is better than the {\tt read}
implementation which only achieves 145~GB/s. This
is explained by the compiler adding loop unrolling and software
prefetching in the assembly code in the {\tt read\_cpp} case, which is
not added in the {\tt read} case. On the other hand, the {\tt write}
case reaches a performance of 158~GB/s while the
{\tt write\_cpp} implementation never exceeds 58~GB/s. 
This difference is due to the use of the {\tt \_mm512\_store\_nrngo\_ps} instruction, 
which bypasses the Read For Ownership protocol and allows the writes
to be performed in any order. The compiler can not use this
instruction on its own because this would break the memory convention
useful for some parallel algorithms. The {\tt read/write} benchmark
performs at 140~GB/s while the {\tt read/write cpp} benchmark performs
at about 90~GB/s.

Note that SpMV kernels will likely use only simple read and simple
writes on the MIC architecture. We also benchmark the performance we
can achieve using {\tt gather} and {\tt unpack} operations and we
present the results in Figure~\ref{fig:band_gather}. The {\tt unpack}
instruction can reach a performance of 140~GB/s. (There is no easy way
to implement the equivalent of the {\tt unpack} operation without
using explicit registers.) The performance of the {\tt gather}
operations depends in which order the array is read. If the indices
are ordered ({\tt compact}) then a performance of 140~GB/s is
reached. If the indices are completely randomized, the performance
drops below 5~GB/s. We can see that implementing the
indirection mechanism without vector instructions lowers the
performance significantly. (In the {\tt compact} case, it never
surpasses 100~GB/s and is close to 50~GB/s most of the time.)

@@

To summarize, using vector instructions appears necessary in order
toreach the highest bandwidth on the MIC architecture. For the instructions used
in a typical SpMV computation, one can expect a bandwidth between
140~GB/s to 160~GB/s. That is why we use in our estimations a best
achievable bandwidth of 150~GB/s despite the fact that a 
carefully crafted code to maximize bandwidth can reach higher
values~\cite{Saule13-ARXIV}.

@@

The different matrices used to perform our experiments and analysis
are presented in Figure~\ref{fig:spy_plots}. The
Supercompact matrices (Figure~\ref{fig:supercompact}) have been
generated to only have nonzero elements in the first 32 columns of
the matrix. As a result, only 32 values of {\tt x} are read
from memory, and the expense associated with cache misses and gather
operations is removed from consideration. The cost associated with 
{\tt A} remains.   \todo{reread the above. Reread by Gordon}
The Compact matrices
(Figure~\ref{fig:compact}) are generated to have 32 nonzeros per row
centered around the diagonal and it represents the ideal case for many
applications that rely on sparse matrix vector multiplication. The nonzero
elements of the Random
matrices (Figure~\ref{fig:compact}) see their nonzero elements
randomly (uniformly) distributed in the matrix; they represent the
worst case scenario for a cache based architecture where the cache
reutilisation is the lowest from one row to the next.

The other two types of matrices are used in derivative stencils in 3D
RBF-FD calculations (Figure~\ref{fig:rbf3dnorcm} shows a $32$-points
stencil of a 3D $8^3$ grid). We also apply a Cuthill-Mckee reordering to
all the matrices.  This ordering technique is designed to reduce the
distance between the nonzeros and the diagonal, hopefully allowing a
better cache reutilisation. The reordered version of the matrix can be
seen in Figure~\ref{fig:rbf3drcm}.

@@

  \caption{Performance of the manually vectorized code. The two
    horizontal lines depict the worst and the best predicted performances. \todo{font size}}
  \label{fig:expe_types}
\end{figure*}

We now investigate the actual performance that we can obtain when
multiplying four vectors by four matrices in single
precision. Figure~\ref{fig:expe_types} presents results for different
types of matrices and also shows the minimum and maximum performance
predicted in Section~\ref{sec:model}. Figure~\ref{fig:perf_ideal}
gives the results for the {\tt supercompact} and {\tt random} matrices
that represent both the best and the worst case for such a practical 
computation. The {\tt supercompact} case peaks at
208~Gflop/s being very close to the predicted peak
performance of 213~Gflop/s. Conversely the performance of the {\tt
  random} case decreases to 56~Gflop/s which is close to
the lowest predicted performance of 55~Gflop/s. One can see that RCM
ordering helps the {\tt random} case but the impact decreases when the
size of the matrix increases.

Figure~\ref{fig:perf_realistic} gives the results on the {\tt compact}
case, which represents the best realistic matrix one could find with very
structured grids and a
{\tt RBF} derivative matrix extracted from a real 3D application. In the {\tt
  compact} case, the performance can be as high as 195~Gflop/s, which is
within 15\% of the peak predicted performance. The performance of the
{\tt RBF} case varies between 100~Gflop/s and 140~Gflop/s. Most of the
time, the RCM ordering provides an improvement which can be as high as
30~Gflop/s.

@@

We finally investigate the central questions of this paper. Do we gain
actual performance by transforming classical SpMV computation into the
multiplication of four vectors by four matrices for the computation of the
derivative of RBFs? Does using manual vectorization improve actual
application performance? Figure~\ref{fig:comparison_spmv} compares the
performance achieved by a classical SpMV and by the multiplication of
four vectors by four matrices using either standard C++ code and our
optimized implementation. The
classical SpMV computation reaches 14~Gflop/s. 
The standard C++ implementation reaches a performance of 38~Gflop/s while our
optimized implementation reaches 140~Gflop/s. 
Notice that \todo{I do not know that I ran host and mic on the same grid, but your 
numbers are approximately correct. The numbers are a function of resolution} 
RCM ordering has no impact in the standard C++ implementation while it
provides a significant improvement in our manually vectorized
version. This highlights that the C++ implementation is instruction-bound
while our manually vectorized implementation is most likely 
memory-bound.

To put things in perspective, we conducted some experiments on a
classical CPU architecture with two Intel Core i7 processors, each has 8
cores that support AVX instructions. Both the compiler-generated and
the manually-written vectorization using AVX instruction reach a
(read) bandwidth of 70~GB/s. This indicates that the performance of
the classical SpMV can not overcome 17~Gflop/s. A generic
implementation of the multiplication of four vectors with four matrices on
the {\tt RBF} matrix reaches a performance of 27~Gflop/s; while a
manual implementation using AVX instruction reaches a performance of
42~Gflop/s. This confirms that our method is sound and generalizes to
architectures than the Intel MIC architecture. \todo{albeit with less 
impressive results given that swizzling is often not available. Swizzling
will become available in a new version of the host: enhanced AVX.}

>>>>

\begin{document}
\title{Acceleration of Derivative Calculations with Application to Radial Basis Function --
Finite-Differences on the Intel MIC Architecture}
%\title{Sparse Matrix Vector Multiplication with Multiple vectors and
  %Multiple Matrices on the MIC Architecture \todo{? more application oriented''}}

@@

In this paper, we consider the case of local stencils with the number
of nodes at each point and encode the sparse matrix in ELLPACK
format. We increase the number of operations relative to memory
bandwidth by calculating four derivatives of four different functions,
or 16 different derivatives. We demonstrate a novel implementation on
the Intel MIC architecture, taking into account its advanced swizzling and
channel interchange features. We present benchmarks on a real data
set that show an almost eightfold increase in speed compared to
efficient implementations of a single derivative, reaching a
performance of almost 140~Gflop/s. We explain the results through
consideration of operation count versus memory bandwidth.
\end{abstract}

@@

We discuss how the implementation of the v4m4 kernel works using
vectorial instructions. The code is given for reference in
Figure~\ref{code:mat_mul}. (We will show in Section~\ref{sec:expe}
that an implementation that relies on compiler vectorization does not lead
to desirable performance, which confirms the results
of~\cite{Saule13-ARXIV}.) The registers in the MIC architecture are
512 bits wide and can store 16 floats or integers. (In comparison to
OpenCL or CUDA, one can think of these registers as a warp.) The end
goal is to load and format the nonzero values $A_h^1,A_h^2,A_h^3,A_h^4$
and the vector entries $x_m^1,x_m^2,x_m^3,x_m^4$ in two vector registers and 
apply Fused Multiply-Add on them, which is depicted in the {\tt accu} line of 
Figure~\ref{code:mat_mul}. Bringing the data into the vector
register as depicted in the figure is the difficult component of our 
proposed algorithm, which we proceed to explain further. 
A thread will perform the multiplications one row at a time,
and parallelism is achieved by giving blocks of rows to each thread
using an OpenMP construct.

The core of the technique is the use of swizzling and permutation
features of the vectors in the Intel MIC architecture. A vector of 512
bits is composed of four lanes (sometimes called channels) of 128
bits, which are made of four segments of 32 bits. A swizzling
operation allows the reorder of the elements within each lane (see how
{\tt mat} is extracted from {\tt mat\_ent} in
Figure~\ref{code:mat_mul}). While a permutation reorders entire lanes,
without changing their contents (see for instance how {\tt vect} is
generated from {\tt all\_vect} in Figure~\ref{code:mat_mul}).  Both
permutation and swizzling support common reordering and broadcast. In
the code, the ``permute'' function needs to use the {\tt
  \_mm512\_permute4f128\_epi32} instruction which is only defined on
integers and therefore in place typecasts are necessary to convert to
a vector to/from the integer type.  (Note that we prefer swizzling to
permutation if possible because we believe it is cheaper. Many
instructions have swizzling capability, while permuation is its own
instruction.

Every operation that loads data from memory into a register
transfers 512 bits, or 16 floats at a time. A single nonzero
in the matrix will cause 16 operations to be performed, but it only
uses 4 floats from the matrix and 4 floats from the {\tt x} vector. So
we would like to execute a single instruction to load the data for four 
nonzeros in the matrix, and another instruction to load the data for
four nonzeros from the vectors.  
(The number of nonzeros on a row in our application is
always a multiple of 4; if it weren't, we would add some explicit
zeroes in the data structure for padding.)

@@

Figure~\ref{fig:bandwidth} shows a benchmark that computes the
bandwidth achieved when reading and/or writing large arrays on both
the MIC card and the CPU architecture. The read benchmark performs a
simple sum of the array. The write benchmark sets all the elemnts of
the array to zero, while the read/write benchmarks copies the array into
another one. We also investigate the difference between a straight C++
implementation (compiled with -O3 with proper tagging to inform the
compiler of memory alignments) and a simple implementation with
explicit use of the vector registers.

The {\tt
  read, cpp} implementation reaches a performance of 168~GB/s
 which is better than the {\tt read}
implementation which only achieves 145~GB/s. This
is explained by the compiler adding loop unrolling and software
prefetching in the assembly code in the {\tt read, cpp} case, which is
not added in the {\tt read} case. On the other hand, the {\tt write}
case reaches a performance of 158~GB/s while the
{\tt write, cpp} implementation never exceeds 58~GB/s. 
This difference is due to the use of the {\tt \_mm512\_store\_nrngo\_ps} instruction, 
which bypasses the Read For Ownership protocol and allows the writes
to be performed in any order. The compiler can not use this
instruction on its own because this would break the memory convention
useful for some parallel algorithms. The {\tt read/write} benchmark
performs at 140~GB/s while the {\tt read/write, cpp} benchmark performs
at about 90~GB/s.

Note that SpMV kernels are not likely to use only simple read and simple
writes. We also benchmark the performance we
can achieve using {\tt gather} and {\tt unpack} operations and we
present the results in Figure~\ref{fig:band_gather}. The {\tt unpack}
instruction can reach a performance of 140~GB/s. (There is no easy way
to implement the equivalent of the {\tt unpack} operation without
using explicit registers.) The performance of the {\tt gather}
operations depends in which order the array is read. If the indices
are ordered ({\tt compact}) then a performance of 140~GB/s is
reached. If the indices are completely randomized, the performance
drops below 5~GB/s. We can see that implementing the
indirection mechanism without vector instructions lowers the
performance significantly. (In the {\tt compact} case, it never
surpasses 100~GB/s and is close to 50~GB/s most of the time.)

@@

To summarize, using vector instructions appears necessary in order
toreach the highest bandwidth on the MIC architecture. For the instructions used
in a typical SpMV computation, one can expect a bandwidth between
140~GB/s to 160~GB/s. That is why we use in our estimations a best
achievable bandwidth of 150~GB/s despite the fact that a 
 code carefully crafted to maximize bandwidth can reach higher
values~\cite{Saule13-ARXIV}.

@@

The different matrices used to perform our experiments and analysis
are presented in Figure~\ref{fig:spy_plots}. The
Supercompact matrices (Figure~\ref{fig:supercompact}) have been
generated to only have nonzero elements in the first 32 columns of
the matrix. As a result, only 32 values of {\tt x} are read
from memory, and the expense associated with cache misses is removed from consideration. The cost associated with 
{\tt A} remains. The Compact matrices
(Figure~\ref{fig:compact}) are generated to have 32 nonzeros per row
centered around the diagonal and it represents the ideal case for many
applications that rely on sparse matrix vector multiplication. The nonzero
elements of the Random
matrices (Figure~\ref{fig:compact}) see their nonzero elements
randomly (uniformly) distributed in the matrix; they represent the
worst case scenario for a cache based architecture where the cache
reutilisation is the lowest from one row to the next.
The other type of matrix is used in derivative stencils in 3D
RBF-FD calculations (Figure~\ref{fig:rbf3dnorcm} shows a $32$-points
stencil of a 3D $8^3$ grid).

We also apply a Reverse Cuthill-Mckee reordering to
all the matrices.  This ordering technique aims to reduce the
distance between the nonzeros and the diagonal, hopefully allowing a
better cache reutilisation. The reordered version of the matrix can be
seen in Figure~\ref{fig:rbf3drcm}.

@@

  \caption{Performance of the manually vectorized code. The two
    horizontal lines depict the worst and the best predicted performances.}
  \label{fig:expe_types}
\end{figure*}

We now investigate the actual performance that we can obtain when
multiplying four vectors by four matrices in single
precision. Figure~\ref{fig:expe_types} presents results for different
types of matrices and also shows the minimum and maximum performance
predicted in Section~\ref{sec:model}. Figure~\ref{fig:perf_ideal}
gives the results for the {\tt supercompact} and {\tt random} matrices
that represent both the best and the worst case for such a practical 
computation. The {\tt supercompact} case peaks at
208~Gflop/s which is very close to the predicted peak
performance of 213~Gflop/s. Conversely the performance of the {\tt
  random} case decreases to 56~Gflop/s which is close to
the lowest predicted performance of 55~Gflop/s. One can see that RCM
ordering helps the {\tt random} case but the impact decreases when the
size of the matrix increases.

Figure~\ref{fig:perf_realistic} gives the results on the {\tt compact}
case, which represents the best realistic matrix one could find with very
structured grids and a
{\tt RBF} derivative matrix extracted from a real 3D application. In the {\tt
  compact} case, the performance can be as high as 195~Gflop/s, which is
within 15\% of the predicted  peak performance. The performance of the
{\tt RBF} case varies between 100~Gflop/s and 140~Gflop/s. Most of the
time, the RCM ordering provides an improvement which can be as high as
30~Gflop/s.

@@

We finally investigate the central questions of this paper. Do we gain
actual performance by transforming classical SpMV computation into the
multiplication of four vectors by four matrices for the computation of the
derivative of RBFs? Does using manual vectorization improve actual
application performance? Figure~\ref{fig:comparison_spmv} compares the
performance achieved by a classical SpMV and by the multiplication of
four vectors by four matrices using either standard C++ code and our
optimized implementation. The
classical SpMV computation reaches 14~Gflop/s. 
The standard C++ implementation reaches a performance of 38~Gflop/s while our
optimized implementation almost reaches 140~Gflop/s. 
Notice that 
RCM ordering has no impact in the standard C++ implementation while it
provides a significant improvement in our manually vectorized
version. This highlights that the C++ implementation is instruction-bound
while our manually vectorized implementation is most likely 
memory-bound.

To put things in perspective, we conducted some experiments on a
classical CPU architecture with two Intel Core i7 processors, each has 8
cores that support AVX instructions. Both the compiler-generated and
the manually-written vectorization using AVX instruction reach a
(read) bandwidth of 70~GB/s. This indicates that the performance of
the classical SpMV can not overcome 17~Gflop/s. A generic
implementation of the multiplication of four vectors with four matrices on
the {\tt RBF} matrix reaches a performance of 27~Gflop/s; while a
manual implementation using AVX instruction reaches a performance of
42~Gflop/s. This confirms that our method is sound and generalizes to
architectures other than the Intel MIC architecture.
